---
output:  github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<center> <h1>HFS-Documentation</h1> </center>
  
<center> <h3><u>table of contents</u></h3> </center>

* [Research Question](#research-questions)

* [Data Cleaning](#data-cleaning)

* [R Script](#r-script)

* [R plots](#r-plots)

* [Results](#results)

* [Conclusion](#ronclusion)

* [Reference](#reference)

## Team:

  * Sai Krishna Gaduputi Subbammagari
  
  * Chad Crowe
  
  * Rhonda Silva

# Licence
[GNU General Public License v3.0](https://github.com/saikrishnags05/Project-for-Data-to-Decisions/blob/master/LICENSE)

# Introduction

Heartland Family Service, which was founded in Omaha in 1875, served more than 79,000 individuals and families last year through direct services, education, and outreach from more than 15 facilities in east central Nebraska and southwest Iowa. In the following focus areas, our programs provide important human services to children, individuals, and families:

•	Housing, Safety, and Financial Stability 

•	Child & Family Well-Being 

•	Counseling & Prevention

Staff can hand-craft an integrated, multi-service, trauma-informed strategy to help clients toward safety, well-being, and, ultimately, self-sufficiency, thanks to their programs and services. A sliding pricing scale is available for some of their counseling services.

Heartland Family Service's objective is to enhance communities by providing education, counseling, and support services to individuals and families. Last year, Heartland Family Service, which was founded in Omaha in 1875, served 60,309 individuals and families through direct services, education, and outreach from more than 15 locations in east central Nebraska and southwest Iowa. In the following target areas:
Child & Family Well-Being, Counseling & Prevention, and Housing, Safety, & Financial Stability,their programs provide important human services to the individuals and families that ultimately create the future of our community.

# Research Questions 

**1)Do significant delays exist between enrolling in an event, entering into the system, and the enrollment approval?**

**Reason:**

It will help the HFS to know the average time taken for a person to complete the full process for any event that is organized in different facility till date. So that If they find any delay in enrolling then they can take action on the particular location and improve the approving process .

**2)Do HFS facility locations have a significant effect on the number of missed appointments?**

**Reason:**

There might be unknown but significant reasons why particular locations have higher rates of missing appointments, such as issues with facility accessibility or local financial burdens like access to transportation.

The second research  question concerns exploring the job role of therapists within HFS, with specific interest in appointment no shows. When patients fail to appear for appointments, this costs HFS time and costs the patient opportunity for therapy. We explore whether there exist clear patterns that might contribute to patients missing visits, such as a location or ethnicity effect. It might be the case that particular facilities are less friendly in supporting a language, which might effect the rate of dropped appointments. 

The research also explores whether job title effects dropped appointments. Job requirements might change from title to title that might have an effect on dropped appointments. This research explores the phenomenon. 

Initial research also explores appointment duration.  Based on the given data, it is unknown whether appointment duration is fixed by insurance or varies between patients. This research explores duration of appointments across job position, ethnicity, location, and the rate of dropped appointments too. While success is not determined by duration, Dr. Juarez mentioned how HFS is very interested in exploring patterns pertaining to the number of appointments and durations by each patient since it affects the funding HFS receives.


**3)Which facilities provide services to  clients identifying as Latino?**

**Reason:**

Which facilities provide services to Latino ethnicity? Would the Clients prefer to have communication in Spanish?  Do these facilities provide Spanish-speaking therapists?

# Data Cleaning 

We have created a new dataset based which can help use to achive a bvest results for our Research Question. Attributes are mentioned bellow.

`facility`, `actual_date`, `event_name`, `date_entered`, `approved_date`, `program_unit_description`, `zip`, `state`,`ethnic_identity`, `Job Title (Therapists I, II, and III)`,`Appointment Duration`.
`Appointment No Shows`.

**step 1:-** collect all the  columns that are required and store them in a data frame **HFS_data**

```{r ,echo=FALSE, include=FALSE}
library('dplyr')
library('ggplot2')
library('plyr')
library('tidyverse')
library('moderndive')

HFS_data<-read.csv("HFS Service Data.csv") # read data set
selected_columns<-c("program_name","facility","actual_date","event_name","date_entered","approved_date","zip","state","age","ethnic_identity","is_noshow", "job_title",'age','program_type',"program_unit_description")

HFS_data<-HFS_data %>%
  select(selected_columns) 
```

 **step 2:-** Since we are using Data format so we have to find the original date. So, we have used as.Date() function to get the actual date format.
 
```{r,echo=FALSE, include=FALSE}
HFS_data<-HFS_data %>%
select(selected_columns)  # this command is used for selecting the required columns
HFS_data$AD<-as.Date(-(HFS_data$actual_date), origin = '2021-08-25')
HFS_data$ED<-as.Date(-(HFS_data$date_entered), origin = '2021-08-25')
HFS_data$AD_year<-(format(as.Date(HFS_data$ED), "%Y"))
```
 
 **step 3:-** Apply step2 to all the date column to verify it in next phase.
 
 **step 4:-** Now create a new column . Now apply add and subtract methods on the dates if we get the values in negatives then it is enrolled before the event,if 0 then they resisted at the moment and if greater the 0 then it is  after the event.
```{r,echo=FALSE, include=FALSE}
HFS_data$AD_ED<-(HFS_data$actual_date-HFS_data$date_entered)
HFS_data$ED_APD<-(HFS_data$date_entered-HFS_data$approved_date)
HFS_data$AD_APD<-(abs(HFS_data$actual_date-HFS_data$approved_date))
HFS_data<-na.omit(HFS_data)
```
 
 **step 5:-** Total we have 5 states in the data set which is mentioned in short form

 **step 6:-** Every State in short form like `IA`,`NE`,`CO`,`NC`,`SC`and later updated to full form of user understanding `iowa`, `Nebrska`,`colorado`, `north carolina`,`south carolina` and we can observe many different plots based on state and there zip code.
```{r,echo=FALSE, include=FALSE}
HFS_data$state<- as.character(HFS_data$state)
HFS_data$state[HFS_data$state == "NE"] <- "nebrska"
HFS_data$state[HFS_data$state == "IA"] <- "iowa"
HFS_data$state[HFS_data$state == "SC"] <- "south carolina"
HFS_data$state[HFS_data$state == "NC"] <- "north carolina"
HFS_data$state[HFS_data$state == "CO"] <- "colorado"
```

**step 7:-** 
The data contains `r nrow(data)` rows. If we filter out NA values for job title there are `r nrow(as_tibble(HFS_data) %>% drop_na(job_title))`. This means each row has a job title and there are no NA values. Given that there is no missing data, there is no need to handle missing data.

**step 8:-**

Most job titles have fewer than fifty instances. Job titles with many instances include therapist, clinical supervisor, case managers, and admin assists. Of those job titles, there are five types of therapists. Given most of the primary job titles are therapists, the exploration of job titles will focus on therapists. We filter the job titles to the various therapist job positions.

```{r,echo=FALSE, include=FALSE}
library('dplyr')
library('tidyverse')
library('moderndive')
#data <- read.csv("/Users/ccrowe/github/isqa8600_ChadCrowe/programs/data/HFS Service Data.csv")
data <- read.csv("HFS Service Data.csv")
tibble_data <- as_tibble(data)
# data header
#head(tibble_data$job_title)
job_title_counts <- tibble_data %>% group_by(job_title) 
```

```{r ,echo=FALSE, include=FALSE}
therapists = data %>% filter(data$job_title == "THERAPIST I" | data$job_title == "THERAPIST II" | job_title == "THERAPIST III" | job_title == "LEAD THERAPIST" | job_title == "Therapist")
```

If we filter out therapists there are only `r nrow(therapists)` rows, so 1500 fewer rows. 

* **Ethnicity**

```{r ,echo=FALSE, include=FALSE}
tibble_data <- as_tibble(HFS_data)
ethnicity <- tibble_data %>% group_by(ethnic_identity) 
# data header
head(tibble_data$ethnic_identity)

```
We see that 15% of all rows are no shows. 15% seems like a surprisingly high number of appointment no shows for any organization. This metric is worth looking into further. There are no NAs in the column or values we want to filter.

**Step 9**
Delete Rows that have "Not Collected" in variable Ethnic_Identity.

**Step 10**
I then discovered that the variable “ethnic_identity” had the following classes 
*Mexican
*Not Collected
*Not Spanish/Hispanic/Latino
*Other Hispanic or Latino
*Unknown

I omitted the rows that had “Not collected” or “Unknown” since this information will not help with the interpretation of data.

I then changed “Mexican” to “Latino” and “Other Hispanic or Latino” to “Latino”
I changed “Not Spanish/Hispanic/Latino” to “Not Latino”
This leaves the data with 2 classes for the variable titled “ethnic_identity”
	Latino
	Not Latino
```{r,echo=FALSE, include=FALSE}
library('plyr')
HFS_d<-HFS_data
HFS.Cleaned<-HFS_d
HFS.Cleaned$program_unit_description<-HFS_d$program_unit_description
HFS.Cleaned$program_name<-HFS_d$program_name
names(HFS.Cleaned)
HFS.Hispanic<-HFS.Cleaned
HFS.Ethnicity2 = HFS.Cleaned[HFS.Cleaned$ethnic_identity != "Not Collected",]
HFS.Ethnicity2$ethnic_identity[HFS.Ethnicity2$ethnic_identity== "Not Spanish/Hispanic/Latino"]<- "Not Latino"
HFS.Ethnicity2$ethnic_identity[HFS.Ethnicity2$ethnic_identity== "Mexican"]<- "Latino"

HFS.Ethnicity2$ethnic_identity[HFS.Ethnicity2$ethnic_identity== "Other Hispanic or Latino"]<- "Latino"
HFS.Ethnicity2 = HFS.Ethnicity2[HFS.Ethnicity2$ethnic_identity !="Unknown",]
HFS.Hispanic.cleaned<-HFS.Hispanic
HFS.Ethnic_Identity<-HFS.Hispanic.cleaned
HFS.Ethnic_Identity.cleaned = HFS.Ethnic_Identity[HFS.Ethnic_Identity$ethnic_identity  != "Not Collected",]
```


# R Script

## Research Question 1

### Iowa

In the below code i have filtered the original data with the state  name and aggregate the whole data based to get an average time taken for completing the HFS process.

```{r,echo=FALSE, include=FALSE}
HFS_data$state[HFS_data$state == "IA"] <- "iowa"
IA<-subset(HFS_data,HFS_data$state=="iowa")#,HFS_data$program_name=='Mental Health')
ag_ia<-aggregate(AD_APD~facility+AD_year+program_name,IA,mean)
ag_ia$AD_APD<-round(ag_ia$AD_APD,0)
names(ag_ia)[names(ag_ia) == "IA$facility"] <- "facility"     
names(ag_ia)[names(ag_ia) == "IA$AD_year"] <-"AD_year"      
names(ag_ia)[names(ag_ia) == "IA$program_name"] <-"program_name" 
names(ag_ia)[names(ag_ia) == "IA$AD_APD"] <-"AD_APD"
```

In the Below code i have build a model to see if it have better confidence in between the attributes in the data frame or not
```{r,echo=FALSE, include=FALSE}
# Build the model

model <- lm(AD_APD ~facility+AD_year, data = ag_ia)
model
new.ag_ia.avg.time <- data.frame(
  facility = c("Thomas Jefferson High School","Titan Hill Intermediate School", "Heartland Family Service - Heartland Homes")
  ,AD_year=c('2021','2020','2020')
  )

predict(model, newdata = new.ag_ia.avg.time)
predict(model, newdata = new.ag_ia.avg.time, interval = "confidence")
```

tried to get a predict value for over all data for Iowa
```{r,echo=FALSE, include=FALSE}
# 1. Add predictions 
pred.int <- predict(model, interval = "prediction")
mydata <- cbind(ag_ia, pred.int)
```

Now i have created a plot to see how the predicted value is similar with 
```{r,echo=FALSE}
# 2. Regression line + confidence intervals
library("ggplot2")
p <- ggplot(mydata, aes(y=fit,x=AD_APD,color=AD_year ))+ geom_point()+geom_smooth(method = "lm")+facet_wrap(~AD_year,scales = 'free')
```

```{r,echo=FALSE}
# 3. Add prediction intervals
p + geom_line(aes(x = lwr), color = "red", linetype = "dashed")+
  geom_line(aes(x = upr), color = "red", linetype = "dashed")

```
(optional) Now for just verification i am trying to get a Residuals of the data that is present in Iowa
```{r,echo=FALSE}
aggregate_IA<-(aggregate(AD_APD~state+facility+AD_year,IA,mean))
aov_group_IA <- aov(AD_APD~facility, data = aggregate_IA)
summary_group_IA<-summary(aov_group_IA)
summary_group_IA
```

```{r,echo=FALSE, include=FALSE}
HFS_data$state[HFS_data$state == "NE"] <- "nebrska"
NE<-subset(HFS_data,HFS_data$state=="nebrska")#,HFS_data$program_name=='Mental Health')
ag_ne<-aggregate(NE$AD_APD~NE$facility+NE$AD_year+NE$program_name,NE,mean)
ag_ne$`NE$AD_APD`<-round(ag_ne$`NE$AD_APD`,0)
names(ag_ne)[names(ag_ne) == "NE$facility"] <- "facility"     
names(ag_ne)[names(ag_ne) == "NE$AD_year"] <-"AD_year"      
names(ag_ne)[names(ag_ne) == "NE$program_name"] <-"program_name" 
names(ag_ne)[names(ag_ne) == "NE$AD_APD"] <-"AD_APD"
# Build the model
model <- lm(AD_APD ~facility+AD_year, data = ag_ne)
new.ag_ne.avg.time <- data.frame(
  facility = c("Heartland Family Service - Heartland Homes","Sanctuary House", "North Omaha Intergenerational Campus (Service)")
  ,AD_year=c('2021','2020','2020')
)
pre<-predict(model, newdata = new.ag_ne.avg.time)
pre_con<-predict(model, newdata = new.ag_ne.avg.time, interval = "confidence")
```

```{r,echo=FALSE,warning=FALSE}

# 2. Regression line + confidence intervals
library("ggplot2")
pred.int <- predict(model, interval = "prediction")
mydata <- cbind(ag_ne, pred.int)
p <- ggplot(mydata, aes(y=fit,x=AD_APD,color=AD_year ))+ geom_point()+geom_smooth(method = "lm")+facet_wrap(~AD_year,scales = 'free')

# 3. Add prediction intervals
p + geom_line(aes(x = lwr), color = "red", linetype = "dashed")+
  geom_line(aes(x = upr), color = "red", linetype = "dashed")
```

```{r,echo=FALSE, include=FALSE}
aggregate_NE<-(aggregate(AD_APD~state+facility+AD_year,NE,mean))
aov_group_NE <- aov(AD_APD~facility+AD_year, data = aggregate_NE)
summary_group_NE<-summary(aov_group_NE)
summary_group_NE
```

## Research Question 3:
### Bar chart of Ethnicity & Facility

This BarChart shows us that the majority of Latinos served attend the
North Omaha Campus and the Heartland Family Service-Central location.
```{r}
ggplot(HFS.Ethnicity2) +
  aes(x = facility, fill = ethnic_identity)+
  labs(fill = "Ethnicity")+
  geom_bar() +
  scale_fill_hue(direction = 1) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.x = element_text(size = 9L))
```



This BoxPlot identifies that most Latinos receive services for Mental
Health Programs and are between the ages of 18 and 50.
```{r}
ggplot(HFS.Ethnicity2) +
  aes(x = program_unit_description, y = age, fill = ethnic_identity) +
  geom_boxplot(width=0.5, lwd=1) +
  scale_color_hue(direction = 1) +
  ylab("Age") + xlab("Program Unit")+
  labs(fill = "Ethnicity") +
  labs (title = "Ethnicity by Age & Program Unit") +
  coord_flip() +
  theme_minimal()
```


# R plots

# Research Question 1
```{r}

HFS_data$state[HFS_data$state == "IA"] <- "iowa"
IA<-subset(HFS_data,HFS_data$state=="iowa")#,HFS_data$program_name=='Mental Health')
ag_ia<-aggregate(IA$AD_APD~IA$facility+IA$AD_year,IA,mean)
ag_ia$`IA$AD_APD`<-round(ag_ia$`IA$AD_APD`,0)
names(ag_ia)[names(ag_ia) == "IA$facility"] <- "facility"     
names(ag_ia)[names(ag_ia) == "IA$AD_year"] <-"AD_year"      
names(ag_ia)[names(ag_ia) == "IA$program_name"] <-"program_name" 
names(ag_ia)[names(ag_ia) == "IA$AD_APD"] <-"AD_APD"

```

If we observe the Graph we can tell that over all behavior of enrollment process w.r.t the  time taken for per person to enroll for an event from past **`r length(unique(IA$AD_year))` years** in the state of **Iowa**.

* There are few facilities where it is taking more time to enroll for a person compared to the previous year.
* Most of the facilities conduct events on **Mental Health**
* We can notice that in the enrolling time is so less in schools and we can even assume that most of the people who enroll to the course are children.
* In **Substance Use** we can clearly tell that out of 5 facilities only 1 facility have organized for more years and their time of enrollment is also almost equal
* From the plot, I can tell that all the enrollments are been late for the past 2 years there may be multiple reasons.
**Example: -** lockdown because Covid-19 which stopped the process


```{r}
#library('plotly')
p <- ggplot(data=ag_ia, aes(x=AD_APD,y=facility,fill=AD_year)) +
  geom_bar(stat="identity", position=position_dodge())+labs(title = "Average Time taken for a person to enroll for a program") +xlab('Average days for a person to register')+ylab('Facility in Iowa')+ labs(colour = "Years")+ scale_fill_brewer(palette="Blues")
#ggplotly(p)
p
```

```{r}
HFS_data$state[HFS_data$state == "NE"] <- "nebrska"
NE<-subset(HFS_data,HFS_data$state=="nebrska")#,HFS_data$program_name=='Mental Health')

ag_ne<-aggregate(NE$AD_APD~NE$facility+NE$AD_year+NE$program_name,NE,mean)

ag_ne$`NE$AD_APD`<-round(ag_ne$`NE$AD_APD`,0)
names(ag_ne)[names(ag_ne) == "NE$facility"] <- "facility"     
names(ag_ne)[names(ag_ne) == "NE$AD_year"] <-"AD_year"      
names(ag_ne)[names(ag_ne) == "NE$program_name"] <-"program_name" 
names(ag_ne)[names(ag_ne) == "NE$AD_APD"] <-"AD_APD"
```
```{r}
p <- ggplot(data=ag_ne, aes(x=AD_APD,y=facility,fill=AD_year)) +
  geom_bar(stat="identity", position=position_dodge())+labs(title = "Average Time taken for a person to enroll for a program") +xlab('Average days for a person to register')+ylab('Facility in Nebraska')+ labs(colour = "Years")+ scale_fill_brewer(palette="Blues")
p
```


If we observe the Graph we can tell that over all behavior of enrollment process w.r.t the  time taken for per person to enroll for an event from past **`r length(unique(NE$AD_year))` years** in the state of **Nebraska**.

* There are few facilities where it is taking more time to enroll for a person compared to the previous year.
* Most of the facilities conduct events on **Mental Health**
* We can also tell that the entire enrollment process is late in all the regions in Nebraska when it is compared to previous years
* From the plot I can tell that all the enrollments are being late for the past 2 years there may be multiple reasons. 

**Example: -** lockdown because Covid-19 which stopped the process

```{r}
#,out.width = "1200",out.height='800'
HFS_data$state[HFS_data$state == "NE"] <- "nebrska"
NE<-subset(HFS_data,HFS_data$state=="nebrska")#,HFS_data$program_name=='Mental Health')

ag_ne<-aggregate(NE$AD_APD~NE$facility+NE$AD_year+NE$program_name,NE,mean)

ag_ne$`NE$AD_APD`<-round(ag_ne$`NE$AD_APD`,0)
names(ag_ne)[names(ag_ne) == "NE$facility"] <- "facility"     
names(ag_ne)[names(ag_ne) == "NE$AD_year"] <-"AD_year"      
names(ag_ne)[names(ag_ne) == "NE$program_name"] <-"program_name" 
names(ag_ne)[names(ag_ne) == "NE$AD_APD"] <-"AD_APD"
p <- ggplot(data=ag_ne, aes(x=AD_APD,y=facility,fill=AD_year)) +
  geom_bar(stat="identity", position=position_dodge())+labs(title = "Average Time taken for a person to enroll for a program") +xlab('Average days for a person to register')+ylab('Facility in Iowa')+ labs(colour = "Years")+ scale_fill_brewer(palette="Blues")
p
```
# Research Question 2



# Research Question 3
The data cleaning results in a new subset of 5 variables based on the attributes analyzed in the first research question. Important columns within this data cleaning are listed below:
  program_unit_desc ethnic_identity   age facility   program_type

This BoxPlot identifies that most Latinos receive services for Mental Health Programs and are between the ages of 18 and 50.
``` {r}
ggplot(HFS.Ethnicity2) +
  aes(x = program_unit_description, y = age, fill = ethnic_identity) +
  geom_boxplot(width=0.5, lwd=1) +
  scale_color_hue(direction = 1) +
  ylab("Age") + xlab("Program Unit")+
  labs(fill = "Ethnicity") +
  labs (title = "Ethnicity by Age & Program Unit") +
  coord_flip() +
  theme_minimal()
```

##Bar chart of Ethnicity & Facility

This BarChart shows us that the majority of Latinos served attend the North Omaha Campus and the Heartland Family Service-Central location. 

```{r }
p<-ggplot(HFS.Ethnicity2) +
  aes(x = facility, fill = ethnic_identity) +
  geom_bar() +
  scale_fill_hue(direction = 1) +
  coord_flip() +
  theme_minimal() +  theme(axis.title.x = element_text(size = 9L),)
p
```


# Results

# Conclusion

RQ1: Different facilities have significantly different delays. Schools are especially fast but counties are slower.

RQ2: Missed appointments are closely related to the facility. Slight differences in missed appointments by the type of therapist 
(I vs II).

RQ3: HFS could possibly benefit by having the bilingual staff at the Sarpy Office as well so that there are 3 branches throughout the metro area to serve Latino client

# Reference

* [https://www.heartlandfamilyservice.org/](https://www.heartlandfamilyservice.org/)

* [https://addictiontreatmentmagazine.com/rehab/heartland-family-service-omaha/](https://addictiontreatmentmagazine.com/rehab/heartland-family-service-omaha/)

* [https://www.heartlandfamilyservice.org/our-mission/](https://www.heartlandfamilyservice.org/our-mission/)

