---
title: "R_Plot_Assignment"
author: "Rhonda Silva"
date: "11/16/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library('ggplot2')
library('tidyverse')
library('dplyr')
HFS <- read.csv("HFS Service Data.csv")
HFS.Ethnicity2<-HFS [c(2,3,4,24,25,45,49)]
```
## RQ 3
The data cleaning results in a new subset of 5 variables based on the attributes analyzed in the first research question. Important columns within this data cleaning are listed below:
  program_unit_desc ethnic_identity   age facility   program_type

This BoxPlot identifies that most Latinos receive services for Mental Health Programs and are between the ages of 18 and 50.
```{r}
ggplot(HFS.Ethnicity2) +
  aes(x = program_unit_description, y = age, colour = ethnic_identity) +
  geom_boxplot(shape = "circle", 
               fill = "#112446") +
  scale_color_hue(direction = 1) +
  coord_flip() +
  theme_minimal()
```

##Bar chart of Ethnicity & Facility

This BarChart shows us that the majority of Latinos served attend the North Omaha Campus and the Heartland Family Service-Central location. 

```{r Facility, include=FALSE}
p<-ggplot(HFS.Ethnicity2) +
  aes(x = facility, fill = ethnic_identity) +
  geom_bar() +
  scale_fill_hue(direction = 1) +
  coord_flip() +
  theme_minimal() +  theme(axis.title.x = element_text(size = 9L),)
```

```{r}
p
```


