---
title: "Rplots"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## RQ 1

**My Research Question is to tell that how much time of delay it take to takes for enrolling into the event, Entering into system and Approving the enrollment.** `By Sai Krishna`

Reason:- 

* This will help in analyzing the time taking per enrolling from starting to ending w.r.t to State Nebraska and Iowa at their facility.

* This will help the HFS to understand the average time taken for enrollment for person

* If we find any kind any slow processing in any of the facility then we can try to improve by providing required technical support or man power to perform the task done in less time.


```{r}
library('ggplot2') # for sample plot if required
library('dplyr') # to use pipelines '%>%' for data set

HFS_data<-read.csv("HFS Service Data.csv") # read data set


selected_columns<-c("program_name","facility","actual_date","event_name","date_entered",   
                    "approved_date","zip","state","age","ethnic_identity")
# Above list is used for selecting the columns in the dataset 

HFS_data<-HFS_data %>%
  select(selected_columns)  # this command is used for selecting the required columns
HFS_data$AD<-as.Date(-(HFS_data$actual_date), origin = '2021-08-25')
HFS_data$ED<-as.Date(-(HFS_data$date_entered), origin = '2021-08-25')
HFS_data$AD_M_Y <- format(as.Date(HFS_data$AD), "%Y-%m")
HFS_data$AD_year<-(format(as.Date(HFS_data$ED), "%Y"))
HFS_data$AD_ED<-(HFS_data$actual_date-HFS_data$date_entered)
HFS_data$ED_APD<-(HFS_data$date_entered-HFS_data$approved_date)
HFS_data$AD_APD<-(abs(HFS_data$actual_date-HFS_data$approved_date))
HFS_data$state[HFS_data$state == "NE"] <- "nebraska"
HFS_data$state[HFS_data$state == "IA"] <- "iowa"
```


```{r}
IA<-filter(HFS_data,HFS_data$state=="iowa")#,HFS_data$program_name=='Mental Health')
ag_ia<-aggregate(IA$AD_APD~IA$facility+IA$AD_year+IA$program_name,IA,mean)
ag_ia$`IA$AD_APD`<-round(ag_ia$`IA$AD_APD`,0)
```

If we observe the Graph we can tell that over all behavior of enrollment process w.r.t the  time taken for per person to enroll for an event from past **`r length(unique(IA$AD_year))` years** in the state of **Iowa**.


```{r}
p <- ggplot(data = ag_ia, aes(y=`IA$AD_APD`,x=`IA$facility`,color= `IA$AD_year` ))
p + geom_point()+facet_wrap(~`IA$program_name`,scale='free')+
theme(axis.text.x = element_text(angle=90, vjust=1, hjust=1))+
  labs(title = "Average Time taken for a person to enroll for a program") +ylab('Average days for a person to register')+xlab('Facility in Iowa')+ labs(colour = "Years")

```


```{r}
ne<-filter(HFS_data,HFS_data$state=="nebraska")#,HFS_data$program_name=='Mental Health')
ag_ne<-aggregate(ne$AD_APD~ne$facility+ne$AD_year+ne$program_name,ne,mean)
ag_ne$`ne$AD_APD`<-round(ag_ne$`ne$AD_APD`,0)
```

If we observe the Graph we can tell that over all behavior of enrollment process w.r.t the  time taken for per person to enroll for an event from past **`r length(unique(ne$AD_year))` years** in the state of **Nebraska**.

```{r}
#, out.width = "1900px",out.height='800px'
p <- ggplot(data = ag_ne, aes(y=ag_ne$`ne$AD_APD`,x=ag_ne$`ne$facility`,color= ag_ne$`ne$AD_year` ))
p + geom_point()+facet_wrap(~ag_ne$`ne$program_name`,scale='free')+
theme(axis.text.x = element_text(angle=90, vjust=1, hjust=1))+
  labs(title = "Average Time taken for a person to enroll for a program") +ylab('Average days for a person to register')+xlab('Facility in Nebraska')+ labs(colour = "Years")
```
